\addchap{Preface}
\begin{refsection}
 

This collection contains 36 papers presented at the 47th Annual Conference on African Linguistics at UC Berkeley from March 23-March 26, 2016.\footnote{There is one exception: the article entitled `Negation coding in \ili{Ga}' by Yvonne Akwele Amankwaa Ollennu was presented at ACAL44 at Georgetown University but was excluded from that proceedings due to an editorial error.} This meeting of ACAL coincided with a special workshop entitled ``Areal features and linguistic reconstruction in Africa'', and we are glad to include four papers from that workshop in this collection as well. Collectively, these papers add a sizable body of scholarship to the study of African languages, including valuable new descriptions of African languages, novel theoretical analyses of them, and important insights into our typological and historical understanding of these languages. These papers also provide an sample of the depth and richness of contemporary scholarship in linguistics, both in terms of the efficacy of current theories in analyzing language as well as the progress that has been made in describing African languages over the last few decades. Still, much work remains.

Organizing ACAL47 at UC Berkeley was a team effort, and we are grateful to the other members of the organizing committee: Pius Akumbu, Geoff Bacon, Nico Baier, Matthew Faytak, Paula M. Floro, Jevon Heath, Larry Hyman, Maria Khachaturyan, Spencer Lamoureux, Florian Lionnet, John Merrill, and Nicholas Rolle. ACAL47 and the preparation of these proceedings also would not have been possible without the generous support of The National Science Foundation (Conference Grant \#BCS-1546957), the UC Humanities Research Institute, a UC Berkeley Mellon Project Grant, the Doreen B. Townsend Center for the Humanities, the UC Berkeley Center for African Studies, the UC Berkeley Department of Linguistics, and the Association of Contemporary African Linguistics.

For additional technical assistance, help with the conversion of submissions to \LaTeX, and help implementing some of the proofreading changes, we are grateful to Steven Ho, Edwin Ko, Frank Lin, and Tessa Scott, along with the timely assistance and guidance of Sebastian Nordhoff and his team at Language Science Press. In addition, each paper in this volume benefited from the comments of reviewers, who we would like to thank for their work: Pius Akumbu, Felix Ameka, Arto Anttila, Nico Baier, Will Bennett, Lee Bickmore, Adams Bodomo, Gene Buckley, Mike Cahill, Emily Clem, Chris Collins, Denis Creissels, Michael Diercks, Mark Dingemanse, Laura Downing, Phil Duncan, Chris Ehret, Matthew Faytak, Ines Fiedler, Hannah Gibson, Jeff Good, Christopher Green, Lenore Grenoble, Nadine Grimm, Scott Grimm, Fatima Hamlaoui, Brent Henderson, Ken Hiraiwa, Larry Hyman, Peter Jenks, Patrick Jones, Jason Kandybowicz, Maria Khachaturyan, Wendel Kimper, Sampson Korsah, Ruth Kramer, Connie Kutsch-Lojenga, Andrew Lamont, Michael Marlo, Adam McCollum, John Merrill, Irene Monich, Deo Ngonyani, Tatiana Nikitina, Marjorie Pak, Mary Paster, Doris Payne, G\'erard Phillipson, Doug Pulleyblank, Kent Rasmussen, Bert Remijsen, Nicholas Rolle, Sharon Rose, Hannah Sande, Bonnie Sands, Thera Scott, Elisabeth Selkirk, Naga Selvanathan, Ryan Shosted, Harold Torrence, Mauro Tosco, Jochen Trommer, Jenneke van der Wal, Mark Van de Velde, Valentin Vydrin, John Watters, Malte Zimmermann, and Lynell Zogbo. We are especially grateful to those few reviewers who agreed to review more than one paper.

Last, we would like to especially thank the contributors to this volume based in Ghana and Cameroon who were gracious enough not only to attend our conference at UC Berkeley but also to trust their work to us in these proceedings. Their contributions add a depth of perspective that it may be impossible to capture through fieldwork or any class, and we hope that fruitful collaboration between African and Western scholars continues to grow and develop.

\printbibliography[heading=subbibliography]
\end{refsection}

